NanoMaLy

### 2015-11-19 paper/pldi16-submission/haskellListings.tex
down to 11pp
done with abstract

		last,lcm,length,lex,lexDigits,lexLitChar,lines,log,logBase,lookup,map,mapM,mapM_,max,

@@

		readLn,readOct,readParen,readSigned,reads,readsPrec,realToFrac,recip,rem,repeat,replicate,

@@

                len,isBH,isBal,bh,isRB,keys,List,Sorted,RBT,Col,isBlack,OrdRBT,Set,sz,

>>>>

		last,lcm,length,lex,lexDigits,lexLitChar,lines,log,logBase,lookup,mapM,mapM_,max,

@@

		readLn,readOct,readParen,readSigned,reads,readsPrec,realToFrac,recip,rem,repeat,

@@

                len,isBH,isBal,bh,isRB,keys,Sorted,RBT,Col,isBlack,OrdRBT,Set,sz,

### 2015-11-19 paper/pldi16-submission/overview.tex
done with abstract

\section{Overview}
\label{sec:overview}
We start with an overview of our contributions.
%
First we illustrate why generating suitable inputs to functions in the
absense of type information is tricky. 
%
Then we describe our solution to the problem and highlight the
similarity to static type inference.
%
Finally, we demonstrate our visualization of the witnesses we produce.

@@

\paragraph{Problem: Which inputs are bad?}
%
One approach is to randomly generate input values and use them to
execute the program until we find one that causes the program to go
wrong. However, to see why this approach is naive, consider the
@fac@ function from Figure~\ref{fig:factorial}, reproduced here.

@@

To defer input generation, we borrow the notion of a ``hole'' from
SmallCheck~\cite{runciman_smallcheck_2008}.
%
A hole -- written \ehole{} -- is a \emph{placeholder} for a value; in
our case we do not even know its type.
%
We leave all inputs as uninstantiated holes until they are demanded by
the program, \eg due to a primitive operation like the @<=@ test.

@@

As we show in \S~\ref{sec:soundness} our lazy instantiation of holes is
\emph{conservative}.
%
We prove that if calling a function with a hole as input causes the
function to ``go wrong'', then there is \emph{no possible} type for the
function.
%
In other words, for any type you might assign to the function, there
exist valid inputs that will cause it to crash.

\paragraph{Problem: How many inputs does a function take?}
There is another wrinkle though, how did we know that @fac@ takes
a single argument instead of two (or none)?
%
It is clear, syntactically, that @fac@ takes \emph{at least} one
argument, but in a higher-order language with currying syntax can be
deceiving. 
%
Consider the following definition:
%
\begin{code}
  let incAllByOne = List.map (+ 1)
\end{code}
%
Is @incAllByOne@ a function?
%
If so, how many arguments does it take?
%
The \ocaml compiler deduces that @incAllByOne@ takes a single argument
because @List.map@'s \emph{type} says it takes two arguments, and it is
partially applied to @(+ 1)@.
%
As we are dealing with ill-typed programs we do not have the luxury of
typing information, so we must deduce the number of arguments by
experimentation.

\paragraph{Solution: Search for saturated application.}
Luckily the solution is straightforward, we add arguments one-by-one
until we reach a \emph{saturated} application, \ie until evaluating the
application returns a value other than a lambda.

@@

\paragraph{Interacting with the graph}
The reduction graph is useful for formulating and executing traversals,
but displaying it all at once would quickly become overwhelming.
%
Our interaction begins by displaying a big-step reduction, \ie the
witness followed by the stuck term.
%
The user can then progressively fill in the hidden steps of the
computation by selecting a visible term and choosing one of the
applicable traversal strategies -- formalized in
\S~\ref{sec:traversing-graph} -- to insert another term into the
visualization.

>>>>

\section{Overview}
\label{sec:overview}

We start with an overview of our approach to
explaining (static) type errors using \emph{witnesses}
that (dynamically) show how the program goes wrong.
%
We illustrate why generating suitable inputs
to functions is tricky in the absence of type
information.
%
Then we describe our solution to the problem
and highlight the similarity to static type
inference,
%
Finally, we demonstrate our visualization of the witnesses we produce.

@@

\paragraph{Problem: Which inputs are bad?}
%
One approach is to randomly generate input values and
use them to execute the program until we find one that
causes the program to go wrong.
%
Unfortunately, this approach quickly runs aground.
Consider the @fac@ function from Figure~\ref{fig:factorial},
reproduced here.

@@

To defer input generation, we borrow the notion of a ``hole'' from
SmallCheck~\cite{runciman_smallcheck_2008}.
%
A hole --- written \ehole{} --- is a \emph{placeholder} for a value; in
our case we do not even know its type.
%
We leave all inputs as uninstantiated holes until they are demanded by
the program, \eg due to a primitive operation like the @<=@ test.

@@

\paragraph{Witness Generality}
We show in \S~\ref{sec:soundness} our lazy instantiation of holes
produces \emph{general witnesses}.
%
That is, we show that if ``executing''
a function with a hole as input causes the
function to ``go wrong'', then there is
\emph{no possible} type for the function.
%
In other words, for \emph{any} type you might
assign to the function, there exist inhabitants
that will cause the function to go wrong.

\paragraph{Problem: How many inputs does a function take?}
%
There is another wrinkle though, how did we know
that @fac@ takes a single argument instead of two
(or none)?
%
It is clear, syntactically, that @fac@ takes \emph{at least} one
argument, but in a higher-order language with currying syntax can be
deceiving.
%
Consider the following definition:
%
\begin{code}
  let incAllByOne = List.map (+ 1)
\end{code}
%
Is @incAllByOne@ a function?
%
If so, how many arguments does it take?
%
The \ocaml compiler deduces that @incAllByOne@ takes a single argument
because @List.map@'s \emph{type} says it takes two arguments, and it is
partially applied to @(+ 1)@.
%
As we are dealing with ill-typed programs we do not have the luxury of
typing information.

\paragraph{Solution: Search for saturated application.}
We solve this problem by deducing the number of arguments
via an iterative process. We add arguments one-by-one
until we reach a \emph{saturated} application, \ie\
until evaluating the application returns a value
other than a lambda.

@@

\paragraph{Interacting with the graph}
The reduction graph is useful for formulating and executing traversals,
but displaying it all at once would quickly become overwhelming.
%
Our interaction begins by displaying a big-step reduction, \ie the
witness followed by the stuck term.
%
The user can then progressively fill in the hidden steps of the
computation by selecting a visible term and choosing one of the
applicable traversal strategies --- formalized in
\S~\ref{sec:explaining} --- to insert another term into the
visualization.
